Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-11-13 manuscript.Rmd
Update citations & Minor Adjustments

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent class
  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. @vermunt_jk_latent_2004 defined LCA more generally as
virtually any statistical model where "some of the parameters [...]
differ across unobserved subgroups".

@@

In factor analysis, the latent variable is continuous, so the relation
between indicators is define by the communalities between them, and the
latent variable represents the error corrected approximation of the
sample. While in LCA, the relation between indicators is define by how
well they separate the classes, as a multinomial regression between
indicators and latent variable [@masyn_latent_2013]

@@

In this paper, we use the term latent class analysis to refer to a
family of techniques that estimate latent class membership based on a
parametric model of observed indicators. From a historical perspective,
the term was initially conceived to refer to analyses with categorical,
usually binary indicators [@vermunt_jk_latent_2004]. Nowadays, there are
a number of related techniques, known by distinct names, that serve a
similar purpose. The term "latent class analysis" seems most appropriate
as an umbrella term for this broader class of models, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators. Given the
abundance of terms in use for closely related classes of models, we will
provide a rudimentary taxonomy of LCA methods.

A common type of LCA is the *finite Gaussian mixture model*; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes [@collins_latent_2010].

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2010]. As an example, a LPA of ocean
plastic particles found two classes of particles based on length and
width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.
<!-- DA: what is the source for plastic particles? -->

@@

The term latent class analysis originally referred to cases where the
observed indicators were categorical [@collins_latent_2010]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*. Latent class models
with ordinal indicators are parameterized differently from mixture
models. One common parameterization assumes that each categorical
variable reflects an underlying standard normal distribution. The
parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
been distinguished in the literature [@rubin_inference_1976; @enders_applied_2010]: Missing
completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
missing at random (MAR), which means that missingness is contingent on
the *observed* data (and can thus be accounted for); and finally missing
not at random (MNAR), which means that missingness is related to
unobserved factors. It is possible to conduct a so-called "MCAR" test,
for example the non-parametric MCAR test [@jamshidian_tests_2010]. But
note that the name "MCAR test" is somewhat misleading, as the
null-hypothesis of this test is that the data are not MAR, and a
significant test statistic indicates that missingness is related to the
observed data (MAR). A non-significant test statistic does not
distinguish between MCAR or MNAR. As Little's classic MCAR test relies
on the comparison of variances across groups with different patterns of
missing data, it assumes normality [@little_test_1988]. This assumption
is tenuous in the context of LCA. A non-parametric MCAR test, as
provided by Jamshidian and Jalal, may be more suitable
[@jamshidian_tests_2010]. Unfortunately, this test was removed from the
central R-repository CRAN due to lack of maintenance. For this tutorial,
I have re-implemented it in the `mice` package as `mice::mcar()`, with a
fast backend in C++ and new printing and plotting methods.

@@

Conversely, the free open source program OpenMx [@openmx] is very flexible, but
not very user-friendly. We directly address this limitation using the
`tidySEM` R-package. New functionality in `tidySEM` seeks to lower the
threshold for latent class analysis using `OpenMx`. It adheres to best
practices in estimation and reporting, as described in this paper. The
user interface is simple, making use of the model syntax of the widely
used `lavaan` R-package [@lavaan]. This syntax offers a human-readable way to
specify latent variable models. Minor enhancements are made to simplify
the specification of LCA.

>>>>

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent class
  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. @vermunt_latent_2004 defined LCA more generally as
virtually any statistical model where "some of the parameters [...]
differ across unobserved subgroups".

@@

In factor analysis, the latent variable is continuous, so the relation
between indicators is defined by the communalities between them, and the
latent variable represents the error corrected approximation of the
sample. While in LCA, the relation between indicators is define by how
well they separate the classes, as a multinomial regression between
indicators and latent variable [@masyn_latent_2013]

@@

In this paper, we use the term latent class analysis to refer to a
family of techniques that estimate latent class membership based on a
parametric model of observed indicators. From a historical perspective,
the term was initially conceived to refer to analyses with categorical,
usually binary indicators [@vermunt_latent_2004]. Nowadays, there are
a number of related techniques, known by distinct names, that serve a
similar purpose. The term "latent class analysis" seems most appropriate
as an umbrella term for this broader class of models, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators. Given the
abundance of terms in use for closely related classes of models, we will
provide a rudimentary taxonomy of LCA methods.

A common type of LCA is the *finite Gaussian mixture model*; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes [@collins_latent_2009].

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2009]. As an example, a LPA of ocean
plastic particles found two classes of particles based on length and
width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.
<!-- DA: what is the source for plastic particles? -->

@@

The term latent class analysis originally referred to cases where the
observed indicators were categorical [@collins_latent_2009]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*. Latent class models
with ordinal indicators are parameterized differently from mixture
models. One common parameterization assumes that each categorical
variable reflects an underlying standard normal distribution. The
parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
been distinguished in the literature [@rubin_inference_1976; @enders_applied_2022]: Missing
completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
missing at random (MAR), which means that missingness is contingent on
the *observed* data (and can thus be accounted for); and finally missing
not at random (MNAR), which means that missingness is related to
unobserved factors. It is possible to conduct a so-called "MCAR" test,
for example the non-parametric MCAR test [@jamshidian_tests_2010]. But
note that the name "MCAR test" is somewhat misleading, as the
null-hypothesis of this test is that the data are not MAR, and a
significant test statistic indicates that missingness is related to the
observed data (MAR). A non-significant test statistic does not
distinguish between MCAR or MNAR. As Little's classic MCAR test relies
on the comparison of variances across groups with different patterns of
missing data, it assumes normality [@little_test_1988]. This assumption
is tenuous in the context of LCA. A non-parametric MCAR test, as
provided by Jamshidian and Jalal, may be more suitable
[@jamshidian_tests_2010]. Unfortunately, this test was removed from the
central R-repository CRAN due to lack of maintenance. For this tutorial,
I have re-implemented it in the `mice` package as `mice::mcar()`, with a
fast backend in C++ and new printing and plotting methods.

@@

Conversely, the free open source program OpenMx [@neale_openmx_2016] is very flexible, but
not very user-friendly. We directly address this limitation using the
`tidySEM` R-package. New functionality in `tidySEM` seeks to lower the
threshold for latent class analysis using `OpenMx`. It adheres to best
practices in estimation and reporting, as described in this paper. The
user interface is simple, making use of the model syntax of the widely
used `lavaan` R-package [@rosseel_lavaan_2012]. This syntax offers a human-readable way to
specify latent variable models. Minor enhancements are made to simplify
the specification of LCA.
